Double White Dwarf Mergers on Adaptive Meshes\\ I. Methodology and Code Verification

### 2015-08-14 papers/wdmerger_I/paper.tex
Deal with a number of recent comments

We employ a ``sponge'' similar to that described by \citet{maestro3}
for the outer regions of the computational domain. After the
hydrodynamics update, we apply a damping force to the momentum
equation as follows:
\begin{equation}
  (\rho \mathbf{u})^{n+1} \to \frac{(\rho \mathbf{u}^{n+1})}{1 + (\Delta t / \Delta t_S) f_S},
\end{equation} 
where $\Delta t_S$ is a timescale for the sponge to operate on, and
$f_S$ is the damping factor.  We choose it so that that the sponge is
non-operational inside a radius $r_S$ from the origin, and fully
applied at a radius $r_S^\prime \equiv r_S + \Delta r_S$. We then
smooth the sponge out between $r_S$ and $r_S^\prime$:
\begin{equation}
  f_{S} = \begin{dcases} 0 & r < r_S \\ \frac{1}{2}\left(1 - \text{cos}\left[\pi\left(\dfrac{r - r_S}{\Delta r_S}\right)\right]\right) & r_S \leq r < r_S^\prime \\ 1 & r \geq r_S^\prime. \end{dcases}
\end{equation}
For the simulations in this paper we set $r_S$ to be 75\% of the \MarginPar{maybe a plot of the sponge profile?}
distance from the origin to the domain boundaries, and $\Delta r_S$ so
that the sponge extends another 10\% of that distance. We set $\Delta
t_S = 0.01$ s, which is of the same order as the as the CFL timestep
for typical problem setups. We find that this sponge is necessary
especially in the rotating reference frame, as standing instabilities
can create very large velocities in the ambient fluid that drag down
the global timestep by up to an order of magnitude.  However, this
provides another reason to distrust the dynamics of the low-density
fluid outside of the stellar orbits. \MarginPar{I am not sure I understand 
this statement}

@@

We loaded a single star of mass $0.9\ \msolar$ onto the grid at the
origin, and evolved it for 200 seconds. Our diagnostic of choice is
the effective radius of the star, determined by the volume of the grid
that has a density greater than $10^3\ \text{g cm}^{-3}$ (see
\autoref{sec:implementation} for details on this measure). This choice
of density is intended to mark a reasonable outer edge to the star
that is not immediately susceptible to the numerical errors prevalent
near the physical edge of the star.
\autoref{fig:single_star_static_radius} shows our results at various
resolutions.  As expected, the star quickly approaches an equilibrium
size that is different (and in this case larger) than the
one-dimensional model. However, the magnitude of this change becomes
smaller with resolution. The star is only approximately in equilibrium
when the coarse grid of $256^3$ zones has a level of refinement that
jumps by a factor of four. However, even then there is a slight uptick
in the size toward the event, implying that the numerical stability is
not guaranteed for arbitrarily long timescales. This result suggests
that when doing the equilibrium initial models that will former the
basis of later calculations, we should carefully monitor the evolution
of the stars before applying any artificial damping to cause the
merger, to ensure that the merger is due to this applied force and not
the intrinsic numerical instability of the stars. \MarginPar{you never referenced figure 4}

\begin{figure}
  \centering
  \includegraphics[scale=0.45]{plots/single_star_static_ke}
  \caption{Time evolution of the kinetic energy of a $0.9 \msolar$ 
    white dwarf. The lines have the same meaning as in \autoref{fig:single_star_static_radius}.
    \label{fig:single_star_static_ke}}
\end{figure}

@@

To see this effect in action for a stellar simulation, we repeated the
test of \autoref{sec:HSE} with a bulk velocity on the grid. We chose a
velocity of $2.56 \times 10^{8}\ \text{cm s}^{-1}$.  \MarginPar{Approx
  what Mach number is this?}  {For context, this is comparable to the
  orbital velocities of the stars in \autoref{sec:kepler}.  This test
  was inspired by \citet{tasker:2008}, who considered a moving galaxy
  cluster and who obtained a long timescale evolution by using
  periodic boundary conditions, so that the cluster would cross the
  domain multiple times throughout the evolution.  We believe that
  periodic boundary conditions are unrealistic for our type of
  simulation, so we prefer to do one continuous simulation where the
  star does not cross the boundaries.  Since our normal grid was not
  large enough to allow the motion to continue for very long, we
  expanded the domain size by a factor of four, and then included an
  extra refined level around the star to keep the effective resolution
  the same. We started the star off in the lower left corner of the
  domain, and pointed its velocity towards the upper right
  corner. This allowed us to evolve the star for the same length of
  time as for the original test. We note that getting the gravity
  boundary conditions right required us to move the origin of the
  problem at the bulk velocity, so that the multipole moments were
  computed with respect to the stellar center.  \MarginPar{are you
    sure that the single WD left quiet in the {\em corner} of the
    bigger domain is as stable as it was in the earlier test?}

@@

Another that we consider is the energy conservation of the system. Recalling \autoref{sec:gravity_hydro_coupling}, 
there are several different methods of applying the gravitational source term to the hydrodynamics equations. In \castro\ we presently 
have four options, controlled by the parameter {\tt castro.grav\_source\_type}, which we will shortern to {\tt gs} for the present 
discussion. ${\tt gs} = 1$ is the standard cell-centered source term for gravity. ${\tt gs} = 2$ is a slight variation on this scheme 
that determines the value of the energy source term after the momentum source term has been applied; the former approach uses 
the uncorrected momenta in calculating $\rho \mathbf{u} g$, which we have found to be more accurate. ${\tt gs} = 3$ is entirely 
different: after calculating the new momenta, we reset the total energy to be equal to the internal energy plus the 
kinetic energy. This has the virtue of ensuring that there is no conflict due to discretization between the momentum and 
energy equations, and also correctly ensuring that the gravitational force does not directly change the internal energy---and thus the temperature---of the fluid. However, it explicitly sacrifices total energy conservation. ${\tt gs} = 4$ is 
the new conservative method of evaluating the energy source terms at cell faces. The results for the change in energy 
after a single orbit are seen in the first column of \autoref{table:sources}. The first two versions give reasonable and 
similar levels of energy conservation. The third has total energy changes on the order of 100\%, but this itself 
does not have a severe effect on the dynamics because in this scheme the total energy variable is effectively a placeholder 
value of the kinetic energy plus internal energy, rather than being evolved directly. The last scheme is nearly two orders 
of magnitude better in energy conservation, justifying the effort in varying the scheme.

@@

Turning to the conservation properties of the system, we examine as fairly typical cases the equal mass system 
in the inertial frame for energy conservation (\autoref{fig:energy_conservation_equal}), and the unequal mass 
system in the rotating frame for angular momentum conservation (\autoref{fig:angular_momentum_conservation_unequal}). 
Angular momentum is conserved to within a few percent over the 25 orbits, while energy conservation is about 
an order of magnitude better. We note that while this is already a fairly good level of energy conservation, it is not 
nearly as good as the results of \citet{marcello:2012}. This is because we reset the internal energy to a level 
corresponding to our temperature floor when it goes negative, while \citeauthor{marcello:2012} do not reset and 
instead ignore the internal energy if it is negative. The resets impose an artificial floor on our ability to 
conserve energy, but they only happen in low-density regions and do not much affect the large-scale dynamics. 
Meanwhile, angular momentum conservation is not quite as good. This is linked to the decline (or increase) in 
the size of the orbit. This implies that we ought to be careful in concluding that at these moderate resolutions 
we can safely evolve systems for many dozens of orbits; this needs to be verified to ensure that an observed 
inspiral and merger is physically (not numerically) motivated.

@@

We use OpenMP to accelerate the work associated with the boxes owned by each 
MPI task. Originally \castro\ used OpenMP to accelerate individual loops in the 
hydrodynamics routines, such as the piecewise-parabolic edge state reconstruction 
and the conservative flux update. However, there is a significant amount of 
overhead associated with generating a new OpenMP region at each of the many 
different loops in a hydrodynamics algorithm. This makes such a strategy 
sub-optimal for use on many-core processors and GPUs. We have recently switched 
to a tiling approach, where an OpenMP region is generated at the start of 
the hydrodynamics routine and the individual threads separately work on 
different partitions of each box. This results in much less overhead for 
the threading. In general we obtain more efficient simulations than 
could be obtained using MPI only, because there are fewer boxes and thus 
less communication for a given number of processor cores. We are currently 
developing an approach to evaluating the equation of state and nuclear reactions 
(and possibly the entire hydrodynamics algorithm) on GPUs, which will allow us
to take advantage of the significant computational resources embedded in GPUs on 
certain systems.

>>>>

We employ a ``sponge'' similar to that described by \citet{maestro3}
for the outer regions of the computational domain. After the
hydrodynamics update, we apply a damping force to the momentum
equation as follows:
\begin{equation}
  (\rho \mathbf{u})^{n+1} \to \frac{(\rho \mathbf{u}^{n+1})}{1 + (\Delta t / \Delta t_S) f_S},
\end{equation} 
where $\Delta t_S$ is a timescale for the sponge to operate on, and
$f_S$ is the damping factor.  We choose it so that that the sponge is
non-operational inside a radius $r_S$ from the origin, and fully
applied at a radius $r_S^\prime \equiv r_S + \Delta r_S$. We then
smooth the sponge out between $r_S$ and $r_S^\prime$:
\begin{equation}
  f_{S} = \begin{dcases} 0 & r < r_S \\ \frac{1}{2}\left(1 - \text{cos}\left[\pi\left(\dfrac{r - r_S}{\Delta r_S}\right)\right]\right) & r_S \leq r < r_S^\prime \\ 1 & r \geq r_S^\prime. \end{dcases}\label{eq:sponge_frac}
\end{equation}
For the simulations in this paper we set $r_S$ to be 75\% of the 
distance from the origin to the domain boundaries, and $\Delta r_S$ so
that the sponge extends another 10\% of that distance. The resulting profile 
is displayed in Figure \autoref{fig:sponge}. We set $\Delta
t_S = 0.01$ s, which is of the same order as the as the CFL timestep
for typical problem setups. We find that this sponge is necessary
especially in the rotating reference frame, as standing instabilities
can create very large velocities in the ambient fluid that drag down
the global timestep by up to an order of magnitude.  This means, however, 
that while the sponge is applied we should be cautious of imputing any physical 
meaning to what is happening in the low-density gas far from the stars.
\begin{figure}
  \centering
  \includegraphics[scale=0.4]{plots/sponge}
  \caption{Radial profile of the hydrodynamical sponge we apply (\autoref{eq:sponge_frac}). 
           We subtract $f_S$ from unity; the value of $1 - f_S$ indicates what happens 
           to the sponged function after the sponge is applied. The sponge has no effect 
           in the inner part of the domain, and is fully applied at the outer edge.\label{fig:sponge}}
\end{figure}

@@

We loaded a single star of mass $0.9\ \msolar$ onto the grid at the
origin, and evolved it for 200 seconds. Our diagnostic of choice is
the effective radius of the star, determined by the volume of the grid
that has a density greater than $10^3\ \text{g cm}^{-3}$ (see
\autoref{sec:implementation} for details on this measure). This choice
of density is intended to mark a reasonable outer edge to the star
that is not immediately susceptible to the numerical errors prevalent
near the physical edge of the star.
\autoref{fig:single_star_static_radius} shows our results at various
resolutions.  As expected, the star quickly approaches an equilibrium
size that is different (and in this case larger) than the
one-dimensional model. However, the magnitude of this change becomes
smaller with resolution. The star is only approximately in equilibrium 
by this measure when the coarse grid of $256^3$ zones has a level of 
refinement that jumps by a factor of four. However, even then there is a slight uptick
in the size toward the end, implying that the numerical stability is
not guaranteed for arbitrarily long timescales. For another view, we 
consider the kinetic energy on the grid, in \autoref{fig:single_star_static_ke}. 
This is a more holistic measure that weights the contribution by the density. 
At the end of the simulation the kinetic energy is not lower at the highest 
resolution than at the lower resolutions. This result suggests
that when doing the equilibrium initial models that will former the
basis of later calculations, we should carefully monitor the evolution
of the stars before applying any artificial damping to cause the
merger, to ensure that the merger is due to this applied force and not
the intrinsic numerical instability of the stars.
\begin{figure}
  \centering
  \includegraphics[scale=0.45]{plots/single_star_static_ke}
  \caption{Time evolution of the kinetic energy of a $0.9\, \msolar$ 
    white dwarf. The lines have the same meaning as in \autoref{fig:single_star_static_radius}.
    \label{fig:single_star_static_ke}}
\end{figure}

@@

To see this effect in action for a stellar simulation, we repeated the
test of \autoref{sec:HSE} with a bulk velocity on the grid. We chose a
velocity of $2.56 \times 10^{8}\ \text{cm s}^{-1}$. For context, this is 
comparable to the orbital velocities of the stars in \autoref{sec:kepler}, and the Mach
number is of order unity in the stellar core at this speed.  This test
was inspired by \citet{tasker:2008}, who considered a moving galaxy
cluster and who obtained a long timescale evolution by using
periodic boundary conditions, so that the cluster would cross the
domain multiple times throughout the evolution.  We believe that
periodic boundary conditions are unrealistic for our type of
simulation, so we prefer to do one continuous simulation where the
star does not cross the boundaries.  Since our normal grid was not
large enough to allow the motion to continue for very long, we
expanded the domain size by a factor of four, and then included an
extra refined level around the star to keep the effective resolution
the same. We started the star off in the lower left corner of the
domain, and pointed its velocity towards the upper right
corner. This allowed us to evolve the star for the same length of
time as for the original test. We note that getting the gravity
boundary conditions right required us to move the origin of the
problem at the bulk velocity, so that the multipole moments were
computed with respect to the stellar center.  \MarginPar{are you
  sure that the single WD left quiet in the {\em corner} of the
  bigger domain is as stable as it was in the earlier test?}

@@

Another that we consider is the energy conservation of the
system. Recalling \autoref{sec:gravity_hydro_coupling}, there are
several different methods of applying the gravitational source term to
the hydrodynamics equations. In \castro\ we presently have four
options, controlled by the parameter {\tt castro.grav\_source\_type},
which we will shortern to {\tt gs} for the present discussion. 
${\tt gs} = 1$ and ${\tt gs} = 2$ are variations on the standard 
cell-centered source term for gravity. The difference between them is that 
${\tt gs} = 2$ determines the value of the energy source term after the momentum
source term has been applied, while ${\tt gs} = 1$ uses the uncorrected
momenta in calculating $\rho \mathbf{u} g$. We have found ${\tt gs} = 2$ to be
more accurate. ${\tt gs} = 3$ is entirely different: after calculating
the new momenta, we reset the total energy to be equal to the internal
energy plus the kinetic energy. This has the virtue of ensuring that
there is no conflict due to discretization between the momentum and
energy equations, and also correctly ensuring that the gravitational
force does not directly change the internal energy---and thus the
temperature---of the fluid. However, it explicitly sacrifices total
energy conservation. ${\tt gs} = 4$ is the new conservative method of
evaluating the energy source terms at cell faces. The results for the
change in energy after a single orbit are seen in the first column of
\autoref{table:sources}. The first two versions give reasonable and
similar levels of energy conservation. The third has total energy
changes on the order of 100\%, but this itself does not have a severe
effect on the dynamics because in this scheme the total energy
variable is effectively a placeholder value of the kinetic energy plus
internal energy, rather than being evolved directly. The last scheme
is nearly two orders of magnitude better in energy conservation,
justifying the effort in varying the scheme.

@@

Turning to the conservation properties of the system, we examine as
fairly typical cases the equal mass system in the inertial frame for
energy conservation (\autoref{fig:energy_conservation_equal}), and the
unequal mass system in the rotating frame for angular momentum
conservation (\autoref{fig:angular_momentum_conservation_unequal}).
Angular momentum is conserved to within a few percent over the 25
orbits, while energy conservation is about an order of magnitude
better. We note that while this is already a fairly good level of
energy conservation, it is not nearly as good as the results of
\citet{marcello:2012}. This is because we reset the internal energy to
a level corresponding to our temperature floor when it goes negative,
while \citeauthor{marcello:2012} do not reset and instead ignore the
internal energy if it is negative. The resets impose an artificial
floor on our ability to conserve energy, but they only happen in
low-density regions and do not much affect the large-scale dynamics.
Meanwhile, relative angular momentum conservation is not quite as good 
as relative energy conservation.  This is
linked to the decline (or increase) in the size of the orbit. This
implies that we ought to be careful in concluding that at these
moderate resolutions we can safely evolve systems for many dozens of
orbits; this needs to be verified to ensure that an observed inspiral
and merger is physically (not numerically) motivated.

@@

We use OpenMP to accelerate the work associated with the boxes owned
by each MPI task. Originally \castro\ used OpenMP to accelerate
individual loops in the hydrodynamics routines, such as the
piecewise-parabolic edge state reconstruction and the conservative
flux update. However, there is a significant amount of overhead
associated with generating a new OpenMP region at each of the many
different loops in a hydrodynamics algorithm. This makes such a
strategy sub-optimal for use on many-core processors and GPUs. We have
recently switched to a tiling approach where an OpenMP region is
generated at the start of the hydrodynamics routine and the individual
threads separately work on different partitions of each box. This
results in much less overhead for the threading. In general we obtain
more efficient simulations than could be obtained using MPI only,
because there are fewer boxes and thus less communication for a given
number of processor cores. We are currently developing an approach to
evaluating the equation of state and nuclear reactions (and possibly
the entire hydrodynamics algorithm) on GPUs, which will allow us to
take advantage of the significant computational resources embedded in
GPUs on certain systems.
